---
title: "Guinea Pigs Tooth Growth Analysis"
author: "DataScienceGB"
date: "24/07/2015"
output: pdf_document
source: help(ToothGrowth)
---

##Overview
After a field and clinical experiment, using 10 guinea pigs to observe the impact of Vitamin C on Tooth Growth, there is a need to use Statistical Methods to determine which of the substances used (Vitamin C or Orange Juice) and doses (.5, 1.0 or 2.0 mg) have most impact on pigs tooth growth. 
We'll use Exploratory Data Analysis (EDA) thechnics and T-Tests methods to find out.

##Exploratory Data Analysis

Let's first start by analysing the data provided 

```{r load data}
library(ggplot2)
library(knitr)
data("ToothGrowth")

str(ToothGrowth)
kable(summary(ToothGrowth))

kable(head(ToothGrowth,3))
kable(tail(ToothGrowth,3))

gm=round(mean(ToothGrowth$len),2)
tmp=NULL
tmp=subset(ToothGrowth,ToothGrowth$sup=="VC")
vcm=round(mean(tmp$len),2)
tmp=subset(ToothGrowth,ToothGrowth$sup=="OJ")
ojm=round(mean(tmp$len),2)

print (c("General mean: ",gm))
print (c("VC mean: ", vcm))
print (c("OJ mean: ",ojm))

```

Fig. 1 Basic Data Summary

As seen from previous figure, and as expected, we have 60 observations with three variables each representing: Tooth length, Supplement (OJ=Orange Juice, VC=Vitamin C) and Dose (.5, 1.0 or 2.0 mg). 

Having a look a the mean values, we also foundout that the OJ mean (20.66) is larger than the overall mean (18.81) and larger the the VC mean (16.96).

Let's have a visual representation of our data.

```{r graph data}

p<-ggplot(ToothGrowth,aes(x=factor(dose),y=len,fill=factor(dose))) +
            geom_boxplot() +
            facet_grid(. ~ supp )+        
            scale_x_discrete("Dose in mg") +
            labs(y="Tooth Growth")+          
          theme( strip.background = element_rect(colour = "black", fill = "orange")                               )

print(p)


```
Fig2. Data Visual Representation

By examining previous graph, we can see that for .5 and 1.0 mg doses, Orange Juice offers more tooth growth than  Vitamin C, meanwhile there is no clear difference when reviewing the 2.0 mg dose.

## Hypothesis Tests

We'll test ToothGrowth data using T-Test method to accept or reject our null hypothesis.
For this tests, we'll **assume** that data is not grouped (and should not be since measured pigs cannot belog to both tests groups) and also that **variance is unequal**.

Lets set our **null hypothesis** (Ho) to be the mean of both sets equal to 0, having:
$$ \mu VC  = \mu OJ = 0 $$
 
### First Test
Lets first test the supplement regardless of the dose.

```{r test 1}

vc=subset(ToothGrowth,ToothGrowth$sup=="VC")
oj=subset(ToothGrowth,ToothGrowth$sup=="OJ")
t.test(vc$len,oj$len,mpaired = FALSE, var.equal = FALSE)
```
Fig3. General Test Vitamin C vs Orange Juice.

We new from Fig1. that OJ's mean is larger than VC's mean (20.66 vs 16.96), however, our T Test shows a p-value larger than our $\alpha$ value used, being 6% lager than 5% (we ran our test at a 95 % confidence interval).


### Second Test
Now that we know that there is no clear winner when considering all the doses, lets break it down, and compare dose against dose in each supplement

```{r test 2}

vc=subset(ToothGrowth,ToothGrowth$sup=="VC")
oj=subset(ToothGrowth,ToothGrowth$sup=="OJ")
## .5 mg doses
vc05=subset(vc,vc$dose==.5)
oj05=subset(oj,oj$dose==.5)
## 1 mg doses
vc10=subset(vc,vc$dose==1.0)
oj10=subset(oj,oj$dose==1.0)
## 2 mg doses
vc20=subset(vc,vc$dose==2.0)
oj20=subset(oj,oj$dose==2.0)

#Now lets run the tests
#.5 mg dose test
t.test(vc05$len,oj05$len,mpaired = FALSE, var.equal = FALSE)

#1.0 mg dose test
t.test(vc10$len,oj10$len,mpaired = FALSE, var.equal = FALSE)

#2.0 mg dose test
t.test(vc20$len,oj20$len,mpaired = FALSE, var.equal = FALSE)

```
Fig4. Diffrent doses Test Vitamin C vs Orange Juice.

Our Fig. 1 EDA visualization shows us better results using Orange Juice (OJ) for .5 and 1 mg doses, previous tests confirm it. In both p-value is smaller that our $\alpha$ value and confidence intervals don't cross 0.

For 2.0 mg doses our T Test shows a p-value larger than our $\alpha$ value, being 9% lager than 5% and confidence interval contains 0. In such dose cases data can't confirm which supplement is better than the other.

###One Last Test
Lets try to find if using 1 mg of OJ might get better results than 2 mg of VC

```{r test 3}

vc=subset(ToothGrowth,ToothGrowth$sup=="VC")
oj=subset(ToothGrowth,ToothGrowth$sup=="OJ")
## OJ 1 mg doses

oj10=subset(oj,oj$dose==1.0)
## VC 2 mg doses
vc20=subset(vc,vc$dose==2.0)

#Now lets run the tests

#1.0 mg dose test
t.test(vc20$len,oj10$len,mpaired = FALSE, var.equal = FALSE)

```
Fig.5 1.0 mg of OJ vs 2.0 mg of VC

Due to a p-value larger than our $\alpha$ and a confidence interval containg 0, we cannot conclude thet using 1.0 doses of OJ will get better results than 2.0 mg does of VC.

## Summary
After reviewing standard R ToothGrowth dataset one can conclude that for smaller doses (.5 mg  and 1.0 mg) OJ yields better results than VC.