---
title: "ToothGrowth Analysis"
author: "DataScienceGB"
date: "24/07/2015"
output: pdf_document
source: help(ToothGrowth)
---

##Overview
After a field and clinical experiment, using 10 guinea pigs to observe the impact of Vitamin C on Tooth Growth, there is a need to use Statistical Methods to determine which of the substances used (Vitamin C or Orange Juice) and doses (.5,1 or 2 mg) have most impact on ToothGrowth. 
Exploratory Data Analysys and T Tests methods to find out

##Exploratory Data Analysis

Let's first start by analysing the data provided 

```{r load data}
library(ggplot2)
library(knitr)
data("ToothGrowth")

str(ToothGrowth)
kable(summary(ToothGrowth))

kable(head(ToothGrowth,3))
kable(tail(ToothGrowth,3))

gm=round(mean(ToothGrowth$len),2)
tmp=NULL
tmp=subset(ToothGrowth,ToothGrowth$sup=="VC")
vcm=round(mean(tmp$len),2)
tmp=subset(ToothGrowth,ToothGrowth$sup=="OJ")
ojm=round(mean(tmp$len),2)

print (c("General mean: ",gm))
print (c("VC mean: ", vcm))
print (c("OJ mean: ",ojm))

```

Fig. 1 Basic Data Summary

As seen from previous figure, as expected, we have 60 observations with three variables each representing: Tooth length, Supplement (OJ=Orange Juice, VC=Vitamin C) and Dose (.5,1 or 2 mg). We also foundout that the OJ mean is larger than the overall mean and larger the the VC mean.

Let's have a visual representation of our data.

```{r graph data}

p<-ggplot(ToothGrowth,aes(x=factor(dose),y=len,fill=factor(dose))) +
            geom_boxplot() +
            facet_grid(. ~ supp )+        
            scale_x_discrete("Dose in mg") +
            labs(y="Tooth Growth")+          
          theme( strip.background = element_rect(colour = "black", fill = "orange")                               )

print(p)


```
Fig2. Data Visual Representation

By examining previous graph, we can see that for .5 and 1 mg doses, Orange Juice offers more tooth growth than  Vitamin C, meanwhile there is no clear difference when reviewing the 2mg dose.

## Hypothesis Tests

We'll test ToothGrowth data using T Tests method to accept or reject our Null Hypothesis.
For this tests, we'll assume that data is not grouped (and should not be since measured pigs cannot belog to both tests groups) and also that variance is unequal.
