---
title: "Weather Impact in Health and Economics"
author: "DataScienceGB"
date: "17/06/2015"
output: html_document
---

##Synopsis

Must human activities, varing from leasure such as vacations and picnics to public health and economics  are affected by weather conditions. But, how big is its impact in terms of money and people healthness?

We'll review the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to picture out Weather's impact on humans injuries and fatalities and the economic damage caused to properties and crops.


##Data Processing

Data used for this analysis is taken directly from [Coursera Reproducible Research Assesment Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

We'll be using sqldf package to manipulate data  as long with classical R vector manipulation to  set data ready for our anlysis.
For figures, please be adised that you'll need xtable, ggplot2 and gridExtra packages installed.

```{r LibraryLoad}
library(sqldf)
library(gridExtra)
library(ggplot2)
library(xtable)
```

First we'll load data into a dataframe to work with it. 
The following code download and load the requiered data.  

```{r DataLoad}
## Check for file existance
## if exists then load it else
## if not exists bring it from source, inflate it and load it
## Note that read.csv function, inflates and reads data at the same time


bz2_storm_file="./data/repdata_data_StormData.csv.bz2"

if (file.exists(bz2_storm_file) ) {
  stormdf=read.csv(bz2_storm_file,stringsAsFactors=F)
} else { if (!file.exists(bz2_storm_file)) {
       fileUrl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
       download.file(fileUrl,dest=bz2_storm_file,method="curl")
       stormdf=read.csv(bz2_storm_file,stringsAsFactors=F)     
       }
}

dim(stormdf)

```

NOAA storm database, uses a code to specify economic impact in order of magnitude. Since report value is in different magnitudes, we need first to set all to the same scale using the following conversion table:

 |NOAA Code|Conversion Value  |
 |---------|------------------|
 |    K    |    1,000         |
 |    M    |    1,000,000     |
 |    B    |    1,000,000,000 |

For the sake of easyness, we'll convert non documented codes using a value of 1 (one) and assume lower case codes are equivalent to their upper case codes.

Once data is equalized, we can then multiply the scale by the property or crop damage to obtain its value in USD.


```{r convert damage to USD}
#Convert PROPDMG and CROPDMG to same "scale"(usd)
stormdf[!(stormdf$PROPDMGEXP %in% c("b","B","k","K","m","M")),"PROPDMGEXP"] = "1"
stormdf[stormdf$PROPDMGEXP %in% c("k","K"),"PROPDMGEXP"] = "1000"
stormdf[stormdf$PROPDMGEXP %in% c("m","M"),"PROPDMGEXP"] = "1000000"
stormdf[stormdf$PROPDMGEXP %in% c("b","B"),"PROPDMGEXP"] = "1000000000"

stormdf[!(stormdf$CROPDMGEXP %in% c("b","B","k","K","m","M")),"CROPDMGEXP"] = "1"
stormdf[stormdf$CROPDMGEXP %in% c("k","K"),"CROPDMGEXP"] = "1000"
stormdf[stormdf$CROPDMGEXP %in% c("m","M"),"CROPDMGEXP"] = "1000000"
stormdf[stormdf$CROPDMGEXP %in% c("b","B"),"CROPDMGEXP"] = "1000000000"

#Convert Strings to Numeric values
stormdf$PROPDMGEXP=as.numeric(stormdf$PROPDMGEXP)
stormdf$CROPDMGEXP=as.numeric(stormdf$CROPDMGEXP)


```

Now we'll group and summarize economic and health impact by event type. This will help us to point out to the mayor responsibles for each group.

```{r Summarize Data}

## Summarize Data to prepare for results
## for both health and economics will only use the top 10 events

## Summarize health_impact by event type
## Select only top 10 events
health_grp=sqldf("select EVTYPE,
                  count(*) as number_of_events, 
                  avg(FATALITIES+INJURIES) as impact_by_event,
                  sum(FATALITIES+INJURIES) as total_impact                  
                  from stormdf
                  where (FATALITIES+INJURIES)!='NA'
                  group by EVTYPE
                  order by 4 desc limit 10")


## Summarize conomic impact by event type
## selects only top 10
econ_grp=sqldf("select EVTYPE,
               count(*) as number_of_events,
              avg(
                   (PROPDMG*PROPDMGEXP) +
                   (CROPDMG*CROPDMGEXP)
                  ) as econ_impact_by_event,
              sum(
                   (PROPDMG*PROPDMGEXP) +
                   (CROPDMG*CROPDMGEXP)
                  ) as econ_impact 
                from stormdf
                where (PROPDMG+CROPDMG)!='NA'
                group by EVTYPE
                order by 4 desc limit 10"
               )
#set impact to millions of usd
econ_grp$econ_impact=round(econ_grp$econ_impact/1000000 ,0)
econ_grp$econ_impact_by_event=round(econ_grp$econ_impact_by_event/1000000 ,0)


```

## Results

Now we are going to get our first results, to do it, we'll use the  summary of "Fatalities and Ijuries" by EVTYPE so we can realize the impact of an atmospheric event in human health.

``` {r PublicHealthImapct, results='asis'}
tab1=xtable(health_grp,caption="Figure 1: Weather impact on public health")
print(tab1, type="html")

```


As we can see by our summary data, TORNADOS have the higher impact to public health, their impact almost dobules then the total impact of the rest of the higher impact events. 

**Attention must be placed into HEAT related events  (EXCESSIVE HEAT and HEAT), eventhough their total impact is almost 10 times less than #TORNADOS, they represent the higher impact by event, meaning that each time a HEAT relanted EVTYPE is presented we'll expect almost 6 times more impact than a TORNADO. **



Now lets review weather impact on economics. 

``` {r EconomicImapct, results='asis'}

#econ_grp
tab2=xtable(econ_grp,caption="Figure 2: Weather impact on public economy")
print(tab2, type="html")

```


As we can see by our summary data, FLOOD events have the higher impact on economics. 

**Attention must be placed into HURRICANE/TYPHOON + HURRICANE events, their impact, each time they hit, is eight times bigger than any other event**  


The following graphs give us a clearer view of weather impact in health and economics:
 
```{r Health impact and EconomicsDamage Plots}


p<-ggplot(health_grp,aes(x=reorder(EVTYPE, +total_impact),y=total_impact)) +
            geom_bar(stat="identity",fill=1:10) +
            scale_fill_manual(values=c(1:10)) +
            coord_flip() +        
            theme_bw(base_family = "Arial") +
            theme(panel.background = element_rect(fill = "lightblue"),
                  panel.grid.minor = element_line(linetype = "dotted")) +
            labs(title="Severe Weather Public Health Impact")+
            labs(y="Injuries + Fatalities" ,x="Event Type")

 g<-ggplot(econ_grp,aes(x=reorder(EVTYPE, +econ_impact),y=econ_impact)) +
            geom_bar(stat="identity",fill=1:10) +
            scale_fill_manual(values=c(1:10)) +
            coord_flip() +            
            theme_bw(base_family = "Arial") +
            theme(panel.background = element_rect(fill = "lightblue"),
                  panel.grid.minor = element_line(linetype = "dotted")) +
            labs(title="Severe Weather Economic Impact")+
            labs(y="Properties and Crops Damage (Millions of Dollars)" ,x="Event Type")



grid.arrange(p, g,nrow=2)
grid.rect(.5,.5,width=unit(.99,"npc"), height=unit(0.99,"npc"), 
          gp=gpar(lwd=3, fill=NA, col="blue"))

``` 

Figure 3: Weather impact in public health and economy


## Summary

After analyzing NOAA database, we are shure that severe weather impacts human activities, what must be decided is if prevention efforts should focus on those events with higher total impact numbers or those that casue the higher impact each time they cross with humans.


